Comparison of MCMC algorithms in Stochastic Volatility Models

### 2023-08-25 writeups/research_proposal/propsal.qmd
typos

# 3) Methodology
The aim of this research is to use Simulation Based Calibration (SBC) to validate the inference of posterior samples generated by different MCMC algorithms in the context of Stochastic Volatility models. This will help identify any bias in computation and inconsistencies in model implementation and estimation strategies. Comparison of the SBC results between different computational approaches may help identify any strengths or shortcomings as well as any errors in model specification.

@@

### Hamiltonian Monte Carlo (No-U-Turn Sampler)
The Stan programming language's implementation of Hamiltonian Monte Carlo will be used for this study. Stan's default algorithm, the No-U-Turn Sampler (Hoffman & Gelman 2014), allows for direct sampling of the specified stochastic volatility model. Hamiltonian Monte Carlo allows for sampling of the generative model and can flexibly handle complicated likelihood functions. This approach will use the same priors as specified in the Gaussian mixture approximation. However, Hamiltonian does not sample directly from the analytical form of conjugate or conditionally conjugate distribution, so different priors can be chosen if necessary.

@@

| **1) for n in 1000 iterations:**
|   2) Draw from prior: $\theta^{sim}\sim\pi (\theta)$
|   3) Simulate dataset with 1000 observations: $y^{sim} \sim p(y|\theta^{sim})$
|   4) Draw 999 posterior samples (post warmup) $\{\theta_1,\dots , \theta_{L}\} \sim p(\theta | y^{sim})$
|   5) Compute rank statistics $r = rank(\{\theta_1,\dots , \theta_{L}\}, \theta^{sim})$

@@

Figure 4 are the chi squared statistic histograms results of Hamiltonian Monte Carlo and Gaussian approximation respectively. The vertical black line marks the 95th quantile which represents the critical value for hypothesis testing. The HMC chi squared statistics look like they are sampled from the correct distribution, with 6.2\% of the samples greater than the 95\% critical value (doing this calculation for 5000 iterations leads to 4.8\% of samples in the critical area). Figure 5 shows the majority of chi squared statistics for Gaussian approximation within the critical area, suggesting the estimation strategy is not well calibrated.

![Chi squared statistics for HMC](../../simulation_output/sbc_ncp_ksc_priors_0.999_adapt_delta_premade_datasets_r7_1000_iterations/chi_sq_hist.png){width=70%}

![Chi squared statistics for KSC](../../simulation_output/sbc_cp_ksc_model_cp_dgf_r1/chi_sq_hist.png){width=70%}

>>>>

# 3) Methodology
The aim of this research is to use Simulation Based Calibration (SBC) to validate the inference of posterior samples generated by different MCMC algorithms in the context of stochastic volatility models. This will help identify any bias in computation and inconsistencies in model implementation and estimation strategies. Comparison of the SBC results between different computational approaches may help identify any strengths or shortcomings as well as any errors in model specification.

@@

### Hamiltonian Monte Carlo (No-U-Turn Sampler)
The Stan programming language's implementation of Hamiltonian Monte Carlo will be used for this study. Stan's default algorithm, the No-U-Turn Sampler (Hoffman & Gelman 2014), allows for direct sampling of the specified stochastic volatility model. Hamiltonian Monte Carlo allows for sampling of the generative model and can flexibly handle complicated likelihood functions. This approach will use the same priors as specified in the Gaussian mixture approximation. However, Hamiltonian Monte Carlo does not sample directly from the analytical form of conjugate or conditionally conjugate distribution, so different priors can be chosen if necessary.

@@

| **1) for sim in 1000 iterations:**
|   2) Draw from prior: $\theta^{sim}\sim\pi (\theta)$
|   3) Simulate dataset with 1000 observations: $y^{sim} \sim p(y|\theta^{sim})$
|   4) Draw 999 posterior samples (post warmup) $\{\theta_1,\dots , \theta_{L}\} \sim p(\theta | y^{sim})$
|   5) Compute rank statistics $r = rank(\{\theta_1,\dots , \theta_{L}\}, \theta^{sim})$

@@

Figure 4 and 5 are the chi squared statistic histograms results of Hamiltonian Monte Carlo and Gaussian approximation respectively. The vertical black line marks the 95th quantile which represents the critical value for hypothesis testing. The Hamiltonian Monte Carlo chi squared statistics look like they are sampled from the correct distribution, with 6.2\% of the samples greater than the 95\% critical value (doing this calculation for 5000 iterations leads to 4.8\% of samples in the critical area). Figure 5 shows the majority of chi squared statistics for Gaussian approximation within the critical area, suggesting the estimation strategy is not well calibrated.

![Chi squared statistics for Hamiltonian Monte Carlo](../../simulation_output/sbc_ncp_ksc_priors_0.999_adapt_delta_premade_datasets_r7_1000_iterations/chi_sq_hist.png){width=70%}

![Chi squared statistics for Gaussian approximation](../../simulation_output/sbc_cp_ksc_model_cp_dgf_r1/chi_sq_hist.png){width=70%}
